Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach

### 2014-05-22 manuscripts/manuscript.Rmd
working on marc's edits

---
layout: 12pt,review
linenumbers: true
title: "Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust ecosystem management.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points.  Before a collapse, we do not know
          where the tipping points lie, if they exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  These two sources of
          uncertainty frustrate state-of-the-art parametric approaches
          to decision theory and optimal control.  Here we illustrate
          how a non-parametric approach using a Gaussian process prior
          provides a more flexible representation of this inherent
          uncertainty. We can then embed the Gaussian process in a
          stochastic dynamic programming framework in order to make
          robust management predictions under both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  In keeping
          with previous results, we find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The Gaussian process approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches. We illustrate this using examples
          of simulated single-species dynamics, where the standard model
          selection approach should be most effective, and find that it
          still fails to account for uncertainty appropriately and leads
          to population crashes, while management based on the Gaussian 
          process approach does not. We discuss opportunities to adapt
          this approach to multiple species models in further work.

@@

Decision making under uncertainty is a ubiquitous challenge of
natural resource management and conservation.  Ecological dynamics
are frequently complex and difficult to measure, making uncertainty
in our understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Concerns
about the potential sudden collapse of fisheries and other ecosystems
[@Worm2006; @Scheffer2009] has increasingly emphasized the difficulties of
formulating even qualitatively correct models of the underlying processes.

We develop a novel approach to address these concerns in the context of fisheries; though
the underlying challenges and methods are germane to many other
conservation and resource management problems.  Economic value and
ecological concern have made marine fisheries the crucible for much
of the founding work [@Gordon1954; @Reed1979; @May1979; @Ludwig1982]
on management under uncertainty.  Global declines [@Worm2006]
and the controversy surrounding their interpretation [@Hilborn2007; @Worm2009] have made understanding these
challenges all the more pressing.

@@

Uncertainty enters the decision-making process at many levels:
intrinsic stochasticity in biological processes, measurements, and
implementation of policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1996;
@Sethi2005], parametric uncertainty [_e.g._ @Ludwig1982; @Hilborn1997;
@McAllister1998; @Schapaugh2013], and model or structural uncertainty
[_e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].  Of these,
structural uncertainty is generally the hardest to quantify. Typical approaches
assume a weak notion of model uncertainty in which the correct model
(or reasonable approximation) of the dynamics must be identified from
among a handful of alternative models.  Here we consider an approach
that addresses uncertainty at each of these levels without assuming the
dynamics follow a particular (i.e. parametric) structure.

Often we do not understand the system dynamics well
enough to know if a model provides a good approximation over the relevant
range of states and timescales (criteria that we loosely refer to as
defining the "right" or "true" model.)  
Typical solutions involve either model choice, model averaging, or
introducing yet greater model complexity of which others may be special
cases (model averaging being one such way to construct such a model)
[@Williams2001; @Athanassoglou2012; @Cressie2009].  Even setting aside
other computational and statistical concerns (e.g. [@Cressie2009]), these
approaches do not address our second concern - representing uncertainty
outside the observed data range.

@@

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated.
This extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
algorithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013].

This difficult position of having neither the true model nor data that
covers the full range of possible states is unfortunately the rule more
than the exception. The potential concern of tipping points in ecological
dynamics [@Scheffer2001; @Polasky2011] reflects these concerns -- as
either knowledge of the true model or more complete sampling of the
state space would make it easy to identify if a tipping point existed.
If we do not know but cannot rule out such a possibility, then we
face decision-making under this dual challege of model uncertainty and
incomplete data coverage.

These dual concerns pose a substantial challenge to existing decision-theoretic
approaches [@Brozovic2011].  Because intervention is often too late 
after a tipping point has been crossed (but see @Hughes2013), management
is most often concerned with avoiding potentially catastrophic tipping
points before any data is available at or following a transition that
would more clearly reveal these regime shift dynamics [e.g. @Bestelmeyer2012].

<!--
_Map_
-->

Here we illustrate how a stochastic dynamic programming (SDP) algorithm
[@Mangel1988; @Marescot2013] can be driven by the
predictions from a Bayesian non-parametric (BNP) approach [@Munch2005a].
This provides two distinct advantages compared with contemporary
approaches.  First, using a BNP sidesteps the need for an accurate
model-based description of the system dynamics.  Second, the BNP can
better reflect uncertainty that arises when extrapolating a model outside
of the data on which it was fit.  We illustrate that when the correct
model is not known, this latter feature is crucial to providing a robust
decision-theoretic approach in face of substantial structural uncertainty.

This paper represents the first ecological application of the SDP
decision-making framework without an a priori model of the underlying
dynamics.  In contrast to parametric models which can only reflect
uncertainty in parameter estimates, the BNP approach provides a more
state-space dependent representation of uncertainty.  This permits
a much greater uncertainty far from the observed data than near the
observed data.  These features allow the GPDP approach to find robust
management solutions in face of limited data and without knowledge of
the correct model structure.

@@

The idea that any approach can perform well without either having to
know the model or have particularly good data should immediately draw
suspicion.  The reader must bear in mind that the strength of our approach
comes not from black-box predictive power from such limited information,
but rather, by providing a more honest expression of uncertainty outside
the observed data without sacrificing the predictive capacity near the
observed data. By coupling this more accurate description of what is known
and unknown to the decision-making under uncertainty framework provided
by stochastic dynamic programming, we are able to obtain more robust
management policies than with common parametric modeling approaches.

@@

The nature of decision-making problems provides a compelling and
pragmatic way to compare models: Rather than compare models in terms
of best fit to data, model performance is defined in the concrete terms
of the decision-maker's objectives. Although much argument can be made
over the 'correct' objective function, e.g. how to account for the
social value of fish left in the sea vs. the commercial value of fish
harvested; see @Halpern2013 for further discussion of this issue.  We can
always compare model performance across multiple potential objectives.
The important point is that a decision-maker does not necessarily need
a model that provides the best mechanistic understanding or the best
long-term outcome, but rather the one that idelivers the best management
performance over the period of interest.

@@

### Statement of the optimal control problem

To illustrate the application of the Gaussian process-based dynamic
programming approach (hereafter GPDP) we focus on the classical problem
of determining a harvest policy that will maximize the expected discounted
profit over a fixed time interval [@Reed1979; @Ludwig1982; @Mangel1988].
For simplicity we assume profit is proportional to the realized harvest.  
For comparison, we consider the performance of the GPDP policy against
policies derived under several alternative parametric models.

We focus on a class of one-dimensional models of population dynamics as a
'best-case' scenario for the parametric model performance relative to
the Gaussian process.   We consider this a best-case scenario for the
following reasons: first, the underlying dynamics are always simuated
from one the parametric models in the set, whereas in reality we never
have the "true" model among our candidates.  By choosing one-dimensional
models with few parameters, we limit the chance that poor performance
will be due to our inability to estimate parameters well, something
that becomes a more severe problem for higher-dimensional parametric
models. Further, the parametric models we consider are those most commonly
used in modeling stock-recruitment dynamics or to model sudden transitions
between alternative stable states.

>>>>

---
layout: 12pt,review
linenumbers: true
title: "Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach"
author: 
  - name: "Carl Boettiger"
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Marc Mangel
    affiliation: cstar
  - name: Stephan Munch
    affiliation: noaa
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: marc
    address: |
      Center for Stock Assessment Research, Department of Applied Math
      and Statistics, University of California,
      Mail Stop SOE-2, Santa Cruz, CA 95064, USA and Department of
      Biology, University of Bergen, Bergen, Norway 9020
  - code: noaa
    address: | 
      Southwest Fisheries Science Center, 
      National Oceanic and Atmospheric Administration, 
      110 Shaffer Road, Santa Cruz, CA 95060, USA

abstract: |
          Model uncertainty and limited data coverage are fundamental
          challenges to robust management of human intervention in a 
          natural system.  These challenges
          are acutely highlighted by concerns that many ecological systems
          may contain tipping points, such as Allee population sizes.  
          Before a collapse, we do not know
          where the tipping points lie, if they exist at all.  Hence,
          we know neither a complete model of the system dynamics nor
          do we have access to data in some large region of state-space
          where such a tipping point might exist.  We illustrate
          how a Bayesian Non-Parametric (BNP) approach using a Gaussian Process (GP) prior
          provides a flexible representation of this inherent
          uncertainty. We embed GPs in a
          stochastic dynamic programming framework in order to make
          robust management predictions with both model uncertainty and
          limited data. We conduct a set of simulations to evaluate this
          approach as compared with the standard approach of using model
          selection to choose from a set of candidate models.  We find that model selection erroneously
          favors models without tipping points -- leading to harvest policies
          that guarantee extinction. The GP approach
          performs nearly as well as the true model and significantly
          outperforms standard appraoches. We illustrate this using examples
          of simulated single-species dynamics, where the standard model
          selection approach should be most effective, and find that it
          still fails to account for uncertainty appropriately and leads
          to population crashes, while management based on the GP 
          approach does not, since it does not underestimate the uncertainty
          outside of the observed data.

@@

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Ecological dynamics are frequently complex and difficult to measure,
making uncertainty in our understanding a persistent challenge.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  Over the last
four decades (but stretching even further back in time), marked by  the
seminal contributions of @Clark1976; @Clark2009 and @Walters1978,
dynamic optimization methods, particularly Stochastic Dynamic Programming
(SDP, @Clark2009), have become increasingly important as a means of
understanding how to management human intervention into natural systems.
Concomittantly, there has been increasing recognition of the importance of
Allee population sizes or `tipping points' [@Scheffer2001;
@Polasky2011] that management actions may inadvertently
cause populations to cross.  @Scheffer2009 emphasized the difficulties of
formulating even qualitatively correct models of the underlying processes.

We develop a novel approach to address these concerns in the context of
fisheries; though the underlying challenges and methods are germane to
many other problems of conservation or natural resource exploitation,
such as forestry.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work [@Gordon1954;
@Clark1976; -@Clark2009; @Reed1979; @May1979; @Ludwig1982] on management
under uncertainty.  Global declines [@Worm2006] and the controversy
surrounding their interpretation [@Hilborn2007; @Worm2009] have made
understanding these challenges all the more pressing.

@@

Uncertainty enters the decision-making process at many levels:
intrinsic stochasticity in biological processes, measurements, and
implementation of policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1996;
@Sethi2005], parametric uncertainty [_e.g._ @Ludwig1982; @Hilborn1997;
@McAllister1998; @Schapaugh2013], and model or structural uncertainty
[_e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].  Of these,
structural uncertainty is generally the hardest to quantify. Typical approaches
assume a weak notion of model uncertainty in which the correct model
(or reasonable approximation) of the dynamics must be identified from
among a handful of alternative models, using either model choice, model averaging, or 
introducing yet greater model complexity of which others may be special
cases (model averaging being one such way to construct such a model)
[@Williams2001; @Athanassoglou2012; @Cressie2009].  Even setting aside
other computational and statistical concerns (e.g. @Cressie2009), these
approaches do not address our second concern - representing uncertainty
outside the observed data range.

@@

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated, and models may be abused [@Mangel2001].
Extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b) decision-theoretic
algorithms rely not only on reasonable estimates of the expected
outcomes, but depend on the weights given to all possible outcomes
[_e.g._ @Weitzman2013].

<!--This difficult position of having neither the true model nor data that
covers the full range of possible states is unfortunately the rule more
than the exception. The potential concern of tipping points in ecological
dynamics [@Scheffer2001; @Polasky2011] reflects these concerns -- as
either knowledge of the true model or more complete sampling of the
state space would make it easy to identify if a tipping point existed.
If we do not know but cannot rule out such a possibility, then we
face decision-making under this 
-->

The dual concerns of model uncertainty and
incomplete data coverage pose a substantial challenge to existing decision-theoretic
approaches [@Brozovic2011].  Because intervention is often (but not always, see @Hughes2013) too late after a tipping point has been crossed, management
is most often concerned with avoiding potentially catastrophic tipping
points before any data is available at or following a transition that
would more clearly reveal these regime shift dynamics [e.g. @Bestelmeyer2012].

We illustrate how a stochastic dynamic programming (SDP) algorithm
[@Mangel1988; @Marescot2013] can be driven by the
predictions from a Bayesian Non-Parametric (BNP) model of population dynamics [@Munch2005a].
This provides two distinct advantages compared with contemporary
approaches.  First, using a BNP sidesteps the need for an accurate
model-based description of the system dynamics.  Second, the BNP can
better reflect uncertainty that arises when extrapolating a model outside
of the data on which it was fit.  We illustrate that when the correct
model is not known, this latter feature is crucial to providing a robust
decision-theoretic approach in face of substantial structural uncertainty.

This paper represents the first ecological application of the SDP
decision-making framework without an a priori model of the underlying
dynamics.  In contrast to parametric models which can only reflect
uncertainty in parameter estimates, the BNP approach provides a more
state-space dependent representation of uncertainty.  This permits
a much greater uncertainty far from the observed data than near the
observed data.  These features allow the Gaussian Process Dynamic Programming (GPDP) approach to find robust
management solutions in face of limited data and without knowledge of
the correct model structure.

@@

<!--
The idea that any approach can perform well without either having to
know the model or have particularly good data should immediately draw
suspicion.  The reader must bear in mind that the strength of our approach
comes not from black-box predictive power from such limited information,
but rather, by providing a more honest expression of uncertainty outside
the observed data without sacrificing the predictive capacity near the
observed data. By coupling this more accurate description of what is known
and unknown to the decision-making under uncertainty framework provided
by stochastic dynamic programming, we are able to obtain more robust
management policies than with common parametric modeling approaches.
-->

@@

To illustrate GPDP, we consider the performance of the GPDP policy against
the policies derived under several alternative parametric models [@Reed1979; @Ludwig1982; @Mangel1988]. 
The nature of decision-making problems provides a compelling and
pragmatic way to compare models: Rather than compare models in terms
of best fit to data, we define model performance in the concrete terms
of the decision-maker's objectives.

<!-- Although much argument can be made
over the 'correct' objective function, e.g. how to account for the
social value of fish left in the sea vs. the commercial value of fish
harvested; see @Halpern2013 for further discussion of this issue.  We can
always compare model performance across multiple potential objectives.
The important point is that a decision-maker does not necessarily need
a model that provides the best mechanistic understanding or the best
long-term outcome, but rather the one that idelivers the best management
performance over the period of interest.
-->

@@

We first describe the requirements of dynamic optimization for the management of human intervention in natural resource systems. After that we describe three parametric models for population dynamics and the GP description of population dynamics.

### Requirements of dynamic optimization

Dynamic optimization  requires characterizing the dynamics of a state variable (or variables), a control action, and a value function.
In this paper, we consider only a single state variable, so that the ideas can be described as simply as possible. That is, we focus on a class of one-dimensional models of population dynamics as a
'best-case' scenario for the parametric model performance relative to
the GP.   This a best-case scenario for the parametric models because the underlying dynamics are always simuated
from one the parametric models in the set, whereas in reality we never
have the "true" model among our candidates.  By choosing one-dimensional
models with few parameters, we limit the chance that poor performance
will be due to our inability to estimate parameters well, something
that becomes a more severe problem for higher-dimensional parametric
models. Further, the parametric models we consider are those most commonly
used in modeling stock-recruitment dynamics or to model sudden transitions
between alternative stable states.
